The Poisson approximate likelihood compared to the particle filter

### 2024-07-27 ms.Rnw
correcting bake syntax and various other edits

\documentclass[10pt]{article}
\usepackage{color}
\usepackage{graphicx} % Required for inserting
\usepackage{booktabs} % For better-quality horizontal lines
\usepackage{multirow}
\usepackage{threeparttable} % For table notes
\usepackage{array} % For table formatting
\usepackage{setspace}
\usepackage{float} % add this in your preamble
\usepackage{amsmath}
\usepackage{tabularx}

@@

\usepackage[round]{natbib}
\usepackage{multirow}
\usepackage{array}
\usepackage{makecell}
\usepackage[font={small,it}]{caption}
\usepackage[utf8]{inputenc}
\usepackage{mathtools}
\usepackage{subcaption}
\usepackage{graphicx, fullpage, verbatim, amsmath}
\usepackage{url, amsfonts, amssymb, amsthm,color, enumerate}
\usepackage{amsfonts}
\usepackage{bm}

\doublespacing
% \usepackage[english]{babel}

% Useful packages
\usepackage{graphicx}
\bibliographystyle{apalike}
\usepackage[colorlinks=true,citecolor=blue]{hyperref}

% Using \doublespacing in the preamble 
% changes the text to double-line spacing
\doublespacing

@@

\begin{abstract}
Filtering algorithms are fundamental for inference on partially observed stochastic dynamic systems, since they provide access to the likelihood function and hence enable likelihood-based or Bayesian inference.
A novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for a certain subset of partially observed Markov process models.
\cite{wwr} supported the importance of their contribution by theoretical analysis and numerical results.
A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
This evidence, based on a model and data from a previous scientific study, suggests at face value that researchers confronted with similar models should use PAL rather than previous particle filter methods.
We show that the evidence is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
If PAL and the particle filter are used on this new scale, the superficial advantage of PAL largely disappears.
On simulations where the model is correctly specified, the particle filter outperforms PAL.
If a poorly fitting specification of initial value parameters is ammended, the particle filter also outperforms PAL on the actual data.
\end{abstract}

@@

graphics.off()      # close graphics windows
library(pomp)
library(magrittr)
library(ggplot2)
library(scales)
library(foreach)
library(doParallel)
library(doFuture)
library(dplyr)
library(Rcpp)
library(RcppArmadillo)
# install.packages('RcppDist')
sourceCpp('rotavirus_normq.cpp')
source("choose_n_fits.R")
source("eval_logLik_pomp.R")

@@

bake(file = paste0(out_dir, "mif2_maximization/round_01/mifs_global_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_01 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir_full,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
        beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
        S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
        S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
        S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = design_matrix[i, ]
    ) 
  }
  mifs_global_01
})

bake(file = paste0(out_dir, "mif2_maximization/round_01/el_01.rds"),{
  el_01 <- eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
  el_01
})
@

@@

<<maximization2, echo=F, warning=F, message=F>>=
set.seed(1234)
sd_02 = 2/3

el_01 <- readRDS(paste0(out_dir,"mif2_maximization/round_01/el_01.rds"))
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0(out_dir,"mif2_maximization/round_02/mifs_global_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_02 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_01[[1]],
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                beta11=0.01*sd_02,phi=0.01*sd_02,
                sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      params = starting_values_02[[i]]
    ) 
  }
  mifs_global_02
})

bake(file = paste0(out_dir, "mif2_maximization/round_02/el_02.rds"),{
  el_02 <- eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
  el_02
})
@

@@

<<maximization3, echo=F, warning=F, message=F>>=
set.seed(12345)
sd_03 = (2/3)^2

el_02 <- readRDS(paste0(out_dir, "mif2_maximization/round_02/el_02.rds"))
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0(out_dir, "mif2_maximization/round_03/mifs_global_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  mifs_global_03 <- foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_02[[1]],
      rw.sd = rw_sd(beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
                beta11=0.01*sd_03,phi=0.01*sd_03,
                sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
                S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
                S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
                S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03),
      params = starting_values_03[[i]]
    ) 
  }
  mifs_global_03
})

bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
  el_03 <- eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
  el_03
})
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
el_03 <- readRDS(paste0(out_dir,"mif2_maximization/round_03/el_03.rds")) 
# Can be changed if more rounds needed
est_mle_vec <- el_03[which.max(el_03$logLik),]
@

@@

<<sim-100, echo=F, warning=F, message=F>>=
bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
  ovov_sim_100 <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

    if (!any(sim_data == 0)) {
      ovov_sim_100[[length(ovov_sim_100) + 1]] <- sim_data
      i <- i + 1  
    } else {
      print(paste("Condition not met for iteration", i, "- retrying"))
    }

    if (length(ovov_sim_100) >= sim_rep) break  
  }
  ovov_sim_100
})
@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
  pfilter_obj <- list()
  pfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

    pfilter_ovov_sim_100[i] <-logmeanexp(logLik(pfs))  
  }
  pfilter_ovov_sim_100
})
@

<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
ovov_sim_100 <- readRDS(paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"))
prop <- 420

bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
  palfilter_ovov_sim_100 <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    pal_sim_dat <- ovov_sim_100[[i]][,-1] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
      c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]), 
      gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
      norm_par = c(0.07,initial_guess[6]), 
      t(prop), NP_MIF,1)
    palfilter_ovov_sim_100[i] <- lik_list$log_lik
  }
  palfilter_ovov_sim_100
})

@@

<<pfilter_wwr, echo=FALSE, message=FALSE, warning=FALSE>>=
bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr.rds"),{
  registerDoFuture()
  plan(multicore, workers = ncores)
  foreach(i=1:ncores, .combine=c,
          .options.future=list(seed=998468235L)
  ) %dofuture% {
    pfilter(sir_wwr,Np=NP_MIF,params=params_wwr)
  } -> pfilter_wwr
  pfilter_wwr
})

bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
  pfilter_wwr_cond_lik <- list()
  registerDoFuture()
  plan(multicore, workers = ncores)
  for(k in 1:ncores){
    pfilter_wwr_cond_lik[[k]] <- pfilter_wwr[[k]]@cond.logLik
  }
  pfilter_wwr_cond_lik
})
@

@@

bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
  palfilter_lik_list_wwr <- rotavirus_SMC_qropxi(init_dist,realdat , 9, 
    c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],initial_guess[5]),     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
    norm_par = c(0.07,initial_guess[6]), 
    t(prop), NP_MIF,1)
  palfilter_lik_list_wwr
})
@

@@

PAL provides an approximate filter, whereas the particle filter provides a Monte~Carlo estimate of an exact filter.
Possible explanations for PAL obtaining a higher log-likelihood than a particle filter include:
\begin{enumerate}
\item The particle filter could have high Monte Carlo variance. This results in negative bias in its log-likilihood estimate due to Jensen's inequality, since the particle filter provides an unbiased likelihood estimate.
\item If the model is misspecified, the approximation error in PAL could lead to a higher log-likelihood estimate than that of the actual misspecified model.
\end{enumerate}
Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
%% resulting in estimates of the conditional log-likelihood, $f_{Y
Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
We found that, in this case, the particle filter out-performs PAL (Figure~\ref{fig:PAL_vs_SMC_ovov_100}).
This supports hypothesis 2, above.
Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.

>>>>

\documentclass[10pt]{article}
\usepackage{color}
\usepackage{graphicx} % Required for inserting
\usepackage{booktabs} % For better-quality horizontal lines
\usepackage{multirow}
\usepackage{makecell}
\usepackage{threeparttable} % For table notes
\usepackage{array} % For table formatting
\usepackage{subcaption}
\usepackage{setspace} % For double-line spacing
\usepackage{float} 
\usepackage{tabularx}
\usepackage[round]{natbib}
\usepackage[font={small,it}]{caption}
\usepackage[utf8]{inputenc}
\usepackage{mathtools}
\usepackage{fullpage}
\usepackage{verbatim}
\usepackage{amsfonts,amssymb,amsthm}
\usepackage{url}
\usepackage[colorlinks=true,citecolor=blue]{hyperref}
% Using \doublespacing in the preamble 
% changes the text to double-line spacing
\doublespacing

@@

\bibliographystyle{apalike}

@@

\begin{abstract}
Filtering algorithms are fundamental for inference on partially observed stochastic dynamic systems, since they provide access to the likelihood function and hence enable likelihood-based or Bayesian inference.
A novel Poisson approximate likelihood (PAL) filter was introduced by \cite{wwr}.
PAL employs a Poisson approximation to conditional densities, offering a fast and consistent approximation to the likelihood function for a certain subset of partially observed Markov process models.
\cite{wwr} supported the importance of their contribution by theoretical analysis and numerical results.
A central piece of evidence for PAL is the comparison in Table~1 of \cite{wwr}, which claims a large improvement for PAL over a standard particle filter algorithm.
This evidence, based on a model and data from a previous scientific study, might suggests that researchers confronted with similar models should use PAL rather than previous particle filter methods.
Alternatively, the improvement in likelihood demonstrated over the results presented by \cite{stocks} might indicate an error in that previous work.
We show that the comparison of log-likelihood values made by \cite{wwr} is flawed because the PAL calculations were carried out using a dataset scaled differently from the previous study.
If PAL and the particle filter are used on this new scale, the superficial advantage of PAL largely disappears.
On simulations where the model is correctly specified, the particle filter outperforms PAL.
If a poorly fitting specification of initial value parameters is ammended, the particle filter also outperforms PAL on the actual data.
\end{abstract}

@@

graphics.off()      # close graphics windows
library(pomp)
library(ggplot2)
library(scales)
library(foreach)
library(doParallel)
library(doFuture)
library(dplyr)
library(Rcpp)
library(RcppArmadillo)
sourceCpp('rotavirus_normq.cpp')
source("choose_n_fits.R")
source("eval_logLik_pomp.R")

@@

bake(file = paste0(out_dir, "mif2_maximization/round_01/mifs_global_01.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      sir_full,
      Np = NP_MIF,
      cooling.fraction.50 = 0.5,
      rw.sd = rw_sd(beta1=0.01,beta2=0.01,beta3=0.01,
        beta11=0.01,phi=0.01,sigma_q=0.01,sigma_xi=0.01,
        S10=ivp(0.24),I10=ivp(0.24),R10=ivp(0.24),
        S20=ivp(0.24),I20=ivp(0.24),R20=ivp(0.24),
        S30=ivp(0.24),I30=ivp(0.24),R30=ivp(0.24)),
      cooling.type = "geometric",
      Nmif = NMIF,
      params = design_matrix[i, ]
    ) 
  }
}) -> mifs_global_01

bake(file = paste0(out_dir, "mif2_maximization/round_01/el_01.rds"),{
  eval_logLik_pomp(mifs_global_01, ncores, NP_MIF, 4248930)
}) ->  el_01

@

@@

<<maximization2, echo=F, warning=F, message=F>>=
set.seed(1234)
sd_02 = 2/3
starting_values_02 <- choose_n_fits(el_01, top_n_fits, ncores)

bake(file = paste0(out_dir,"mif2_maximization/round_02/mifs_global_02.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_01[[1]],
      rw.sd = rw_sd(beta1=0.01*sd_02,beta2=0.01*sd_02,beta3=0.01*sd_02,
                beta11=0.01*sd_02,phi=0.01*sd_02,
                sigma_q=0.01*sd_02,sigma_xi=0.01*sd_02,
                S10=ivp(0.24)*sd_02,I10=ivp(0.24)*sd_02,R10=ivp(0.24)*sd_02,
                S20=ivp(0.24)*sd_02,I20=ivp(0.24)*sd_02,R20=ivp(0.24)*sd_02,
                S30=ivp(0.24)*sd_02,I30=ivp(0.24)*sd_02,R30=ivp(0.24)*sd_02),
      params = starting_values_02[[i]]
    ) 
  }
}) ->  mifs_global_02

bake(file = paste0(out_dir, "mif2_maximization/round_02/el_02.rds"),{
  eval_logLik_pomp(mifs_global_02, ncores, NP_MIF, seed=4248930)
}) ->  el_02
@

@@

<<maximization3, echo=F, warning=F, message=F>>=
set.seed(12345)
sd_03 = (2/3)^2
starting_values_03 <- choose_n_fits(el_02, top_n_fits, ncores)

bake(file = paste0(out_dir, "mif2_maximization/round_03/mifs_global_03.rds"),{ 
  registerDoFuture()
  plan(multicore, workers=ncores)
  foreach(i=1:ncores,.options.future=list(seed=652643293)) %dofuture% {
    pomp::mif2(
      mifs_global_02[[1]],
      rw.sd = rw_sd(
        beta1=0.01*sd_03,beta2=0.01*sd_03,beta3=0.01*sd_03,
        beta11=0.01*sd_03,phi=0.01*sd_03,
        sigma_q=0.01*sd_03,sigma_xi=0.01*sd_03,
        S10=ivp(0.24)*sd_03,I10=ivp(0.24)*sd_03,R10=ivp(0.24)*sd_03,
        S20=ivp(0.24)*sd_03,I20=ivp(0.24)*sd_03,R20=ivp(0.24)*sd_03,
        S30=ivp(0.24)*sd_03,I30=ivp(0.24)*sd_03,R30=ivp(0.24)*sd_03
      ),
      params = starting_values_03[[i]]
    ) 
  }
}) -> mifs_global_03

bake(file = paste0(out_dir,"mif2_maximization/round_03/el_03.rds"),{
  eval_logLik_pomp(mifs_global_03, ncores, NP_MIF, seed=4248930)
}) ->  el_03
@

<<estimated-mle, echo=FALSE, message=FALSE, warning=FALSE>>=
# Can be changed if more rounds needed
est_mle_vec <- el_03[which.max(el_03$logLik),]
@

@@

<<sim-100, echo=F, warning=F, message=F>>=
bake(file=paste0(out_dir, "pfilter/simulated_dataset/ovov_sim_100.rds"),{
  ovov_sim_computation <- list()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

    if (!any(sim_data == 0)) {
      ovov_sim_computation[[length(ovov_sim_computation) + 1]] <- sim_data
      i <- i + 1  
    } else {
      print(paste("Condition not met for iteration", i, "- retrying"))
    }

    if (length(ovov_sim_computation) >= sim_rep) break  
  }
  ovov_sim_computation
}) -> ovov_sim_100
@

<<pfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
bake(file=paste0(out_dir, "pfilter/simulated_dataset/pfilter_ovov_sim_100.rds"),{
  pfilter_obj <- list()
  pfilter_ovov_sim_computation <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)

@@

    pfilter_ovov_sim_computation[i] <- logmeanexp(logLik(pfs))  
  }
  pfilter_ovov_sim_computation
}) ->  pfilter_ovov_sim_100
@

<<palfilter_sim_100, echo=FALSE, message=FALSE, warning=FALSE, results='hide'>>=
prop <- 420
init_dist <- c(3876549, 30351, 1315221, 57139612, 871, 302852, 19573727, 2550, 131092)
initial_guess <- c(11.48,0.25,0.35,0.14,0.16,0.021,66.89)

bake(file=paste0(out_dir, "pfilter/simulated_dataset/palfilter_ovov_sim_100.rds"),{
  palfilter_ovov_sim_computation <- c()
  registerDoFuture()
  plan(multicore, workers=ncores)
  for(i in 1:length(ovov_sim_100)){
    pal_sim_dat <- ovov_sim_100[[i]][,-1] |> as.matrix()
    lik_list <- rotavirus_SMC_qropxi(init_dist,pal_sim_dat , 9, 
      c(initial_guess[1],initial_guess[2],initial_guess[3],
        initial_guess[4],initial_guess[5]), 
        gamma_par = c(initial_guess[7],  1/initial_guess[7]
      ), 
      norm_par = c(0.07,initial_guess[6]), 
      t(prop), NP_MIF,1)
    palfilter_ovov_sim_computation[i] <- lik_list$log_lik
  }
  palfilter_ovov_sim_computation
}) -> palfilter_ovov_sim_100

@@

<<pfilter_wwr, echo=FALSE, message=FALSE, warning=FALSE>>=
bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr.rds"),{
  registerDoFuture()
  plan(multicore, workers = ncores)
  foreach(i=1:ncores, .combine=c,
          .options.future=list(seed=998468235L)
  ) %dofuture% {
    pfilter(sir_wwr,Np=NP_MIF,params=params_wwr)
  } 
}) ->  pfilter_wwr

bake(file=paste0(out_dir,"pfilter/original_dataset/pfilter_wwr_cond_lik.rds"),{
  pfilter_wwr_cond_lik_computation <- list()
  registerDoFuture()
  plan(multicore, workers = ncores)
  for(k in 1:ncores){
    pfilter_wwr_cond_lik_computation[[k]] <- pfilter_wwr[[k]]@cond.logLik
  }
  pfilter_wwr_cond_lik_computation
}) -> pfilter_wwr_cond_lik
@

@@

bake(file=paste0(out_dir,"pfilter/original_dataset/palfilter_lik_list_wwr.rds"),{
  rotavirus_SMC_qropxi(
    init_dist,realdat , 9, 
    c(initial_guess[1],initial_guess[2],initial_guess[3],initial_guess[4],
      initial_guess[5]),     gamma_par = c(initial_guess[7],  1/initial_guess[7]), 
    norm_par = c(0.07,initial_guess[6]), 
    t(prop), NP_MIF,1
  )
}) -> palfilter_lik_list_wwr
@

@@

PAL provides an approximate filter, whereas the particle filter provides a Monte~Carlo estimate of an exact filter.
Possible explanations for PAL obtaining a higher log-likelihood than a particle filter include:
\begin{enumerate}
\item The particle filter could have high Monte Carlo variance. This results in negative bias in its log-likilihood estimate due to Jensen's inequality, since the particle filter provides an unbiased likelihood estimate.
\item If the model is misspecified, the approximation error in PAL could lead to a higher log-likelihood estimate than that of the actual misspecified model.
\end{enumerate}
Log-likelihood is a proper scoring rule for forecasts \cite{gneiting2}, and both the particle filter and PAL construct their log-likelihood estimates via a sequence of one-step forecasts.
Therefore, if the model is correctly specified, the approximation error in PAL can only decrease the expected log-likelihood.
We tested this on simulated data, for which the model of \cite{wwr} is correctly specified.
We found that, in this case, the particle filter out-performs PAL (Figure~\ref{fig:PAL_vs_SMC_ovov_100}).
This supports hypothesis 2, above.
Consequently, we looked for model misspecification by comparing the conditional log-likelihood estimates at each time point, as described in the Methods section.
